---
title : Time Series Forecasting on e-Commerce Data
output:
  html_document:
    theme       : cerulean
    toc         : yes
    highlight   : tango
---

***

## Understand Problem Statement

***

Using e-commerce data, build time series models to predict the price of a certain product **for next 4 months** in advance. 

***

## MODEL FLOW

*** 

### Prepare the environment

```{r echo=FALSE}

# Clear the environment
rm(list = ls(all=TRUE))

```

- Load the libraries

```{r message=FALSE}

library(zoo)
library(dplyr)
library(lubridate)
library(TTR)
library(forecast)
library(DMwR)
library(ggplot2)
library(gridExtra)

```

- Load the dataset

```{r}

commerce_data <- readRDS("Data.rds")

```

### Explore the data

```{r, echo=T}

head(commerce_data, 6)

tail(commerce_data, 6)


```

### Summary Statistics

- Study the structure of data

```{r}
str(commerce_data)
```

- Study the statistics about the data

```{r}
summary(commerce_data)
```

> Observations

- SKU has values 1, 2, 3 so it should be a factor variable.
- Same dates are repeating for same SKU also so we will have to use some aggregation function to reduce to price value each day per SKU.
- Some dates are missing while we need all dates in sequence for time series forcasting. Therfore, we will have to impute some value in that place.

### Data Preprocessing 

- Data Type Conversion

```{r, echo=T}

commerce_data$SKU = as.factor(commerce_data$SKU)
commerce_data$Date = ymd(commerce_data$Date)

```

```{r, echo=T}

table(commerce_data$SKU)

```

### Exploratory Data Analysis

Let us extract year, month, day as separate columns using functions from lubridate library.

```{r}

commerce_EDA = commerce_data %>% mutate(year = year(Date), month = month(Date), day = day(Date))
commerce_EDA
```

Let us visualise the price change for each SKU.

```{r}

p1 <- commerce_EDA %>% group_by(SKU, day) %>% summarise(MeanPrice = mean(Price)) %>% 
  ggplot(aes(x = day, y = MeanPrice)) + geom_line() + facet_wrap(~ SKU) +
  theme_minimal()

p2 <- commerce_EDA %>% group_by(SKU, month) %>% summarise(MeanPrice = mean(Price)) %>% 
  ggplot(aes(x = month, y = MeanPrice)) + geom_line() + facet_wrap(~ SKU) + theme_minimal()

p3 <- commerce_EDA %>% group_by(SKU, year) %>% summarise(MeanPrice = mean(Price)) %>% 
  ggplot(aes(x = year, y = MeanPrice)) + geom_line() + facet_wrap(~ SKU) + theme_minimal()

grid.arrange(p1, p2, p3, nrow = 3)

```

We see that the mean price range is very different for each SKU. It is a good idea to build separate models for each SKU. 

- We will build a model for SKU = 3. As we are using only 1 SKU, this column becomes redundant and can be dropped.
- We will use the mean price each date for forecasting.

```{r}

commerce_3 <- commerce_data %>% filter(SKU == 3) %>% dplyr::select(-SKU) %>% 
  group_by(Date) %>% summarise("MeanPrice" = mean(Price))

head(commerce_3, 10)

```

- We will add another variable with all of the dates and join it with our data. This will induce some NA values.

```{r}

dates <- data.frame("Date" = seq(min(commerce_3$Date), max(commerce_3$Date), by="days"))

commerce_3 <- dates %>% full_join(commerce_3, c("Date" = "Date"))

head(commerce_3, 10)

```

- We will replace NA with the average of previous and next values. We will use `na.locf()` function for this. (LOCF = Last Observation Carried Forward)

```{r}

# fromLast = TRUE works reverse
commerce_3$MeanPrice = (na.locf(commerce_3$MeanPrice) + na.locf(commerce_3$MeanPrice, fromLast = T))/2

head(commerce_3, 10)

plot(commerce_3$MeanPrice, type = 'l')

```

We will convert this problem from daily forcasting to monthly forcasting. Lets us extract year and month using lubridate library.

```{r}

commerce_3 <- commerce_3 %>% mutate(Year = year(Date), Month = month(Date))

head(commerce_3)

```

- Drop the Year attribute as it is not required.
- Month attribute ( for use in future model by seasonality ).
- Create sequence Time variable.

```{r}

commerce_3 <- commerce_3 %>% group_by(Year, Month) %>% summarise("MeanPrice" = mean(MeanPrice)) %>% 
  as.data.frame() %>% 
  dplyr::select(-Year) %>% mutate_at('Month', as.factor) %>% mutate(Time = 1:n())

head(commerce_3)

plot(commerce_3$MeanPrice, type = 'l')

```


### Train and Test Split

* As this data set is time dependent and sequence is important i.e. no random split. 

```{r, echo=T}

train = commerce_3[1:53,]
test = commerce_3[54:nrow(commerce_3),]

```

### MODEL BUILDING

#### Regression on time

**As trend is strong Regression on time will work.**

##### Simple Linear Regression

```{r}

lm1 = lm(MeanPrice~Time, data = train)

pred_Train = predict(lm1)
pred_Test  = predict(lm1, test)

plot(train$MeanPrice, type="l")
points(train$Time, pred_Train, type="l", col="red", lwd=2)

```

##### Evaluate SLR Model

```{r, echo=T}
lm1TrainError = regr.eval(train$MeanPrice, pred_Train)
lm1TestError  = regr.eval(test$MeanPrice, pred_Test)
lm1TrainError
lm1TestError
```

##### Quadratic Linear Regression 

```{r}

lm2 = lm(MeanPrice~poly(Time, 2, raw=TRUE), data = train)

pred_Train = predict(lm2)
pred_Test  = predict(lm2, test)

plot(train$MeanPrice, type="l")
points(train$Time, pred_Train, type="l", col="red", lwd=2)

```

##### Evaluate Quadratic LR Model

```{r, echo=T}

lm2TrainError = regr.eval(train$MeanPrice, pred_Train)
lm2TestError  = regr.eval(test$MeanPrice, pred_Test)
lm2TrainError
lm2TestError

```

##### Seasonal Linear Regression  

We will use dummies for Month attribute

```{r}

str(train)
slm1 <- lm(MeanPrice~., data=train)

pred_Train = predict(slm1)
pred_Test  = predict(slm1, test)

plot(train$MeanPrice, type="l")
points(train$Time, pred_Train, type="l", col="red", lwd=2)

```

##### Evaluate Seasonal LR Model 

```{r, echo=T}
slm1TrainError = regr.eval(train$MeanPrice, pred_Train)
slm1TestError = regr.eval(test$MeanPrice, pred_Test)
slm1TrainError
slm1TestError
```

#### Past Value Averaging

- Convert data into R time series object 

```{r, echo=T}

train_TS <- ts(train$MeanPrice, frequency = 12, start = c(2009, 4))
train_TS

test_TS <- ts(test$MeanPrice, frequency = 12, start = c(2013, 9))
test_TS

```

##### Simple Moving Average

```{r, warning=FALSE}

fitsma = SMA(train_TS, n = 2)

# forecast for next 4 time periods
predsma = forecast(fitsma, h = 4)
plot(predsma)

```

- Observe the calculations

```{r}

train_TS
fitsma
predsma$mean
predsma

```

##### Evaluate SMA

```{r, echo=T}

smaTrainError = regr.eval(train_TS[2:length(train_TS)], fitsma[2:length(train_TS)])
smaTrainError

smaTestError = regr.eval(test$MeanPrice, predsma$mean)
smaTestError

```

##### Weighted Moving Average

```{r, echo=T}

fitwma = WMA(train_TS, n=3, 1:3)
predwma = forecast(fitwma, h=4)
plot(predwma)

```

```{r}

train_TS
fitwma
predwma

```

##### Evaluate WMA

```{r, echo=T}

wmaTrainError = regr.eval(train_TS[2:length(train_TS)], fitwma[2:length(train_TS)])
wmaTrainError

wmaTestError = regr.eval(test$MeanPrice, predwma$mean)
wmaTestError

```

##### Exponential Moving Average

```{r, echo=T}

fitEma = EMA(train_TS, n=2)
predema = forecast(fitEma, h=4)
plot(predema)

```

- Observe the calculations

```{r}

train_TS
fitEma
predema

```

##### Evaluate EMA

```{r, echo=T}

emaTrainError = regr.eval(train_TS[2:length(train_TS)], fitEma[2:length(train_TS)])
emaTrainError

emaTestError = regr.eval(test$MeanPrice, predema$mean)
emaTestError

```

#### HoltWinters model

##### Decomposed Time Series

* Decompose will provide more information on seasonality, trend and randomness

```{r, echo=T}

train_Decomposed = decompose(train_TS)
plot(train_Decomposed)

```

##### HoltWinters Additive Model

Since we are building the models on monthly data, you will get 12 seasonal components. If you are reading the weekly data, you will get 53 seasonal components.

```{r}

model_HW_Add = HoltWinters(train_TS, seasonal="additive")
model_HW_Add

```  
  
##### HoltWinters Multiplicative Model

```{r}

model_HW_Mul = HoltWinters(train_TS, seasonal="multiplicative")
model_HW_Mul

```

##### Evaluate HoltWinters Multiplicative Model

```{r}

pred_train_HW = data.frame(model_HW_Mul$fitted)
pred_train_HW

```
  
```{r}

pred_test_HW = forecast(model_HW_Mul, h = 4)

plot(pred_test_HW)

pred_test_HW

regr.eval(test$MeanPrice, pred_test_HW$mean)

```
  
##### HoltWinters Model with no Trend and no Seasonality 

```{r, echo=T}

hw_NT_NS = HoltWinters(train_TS, beta = F, gamma = F)
hw_NT_NS

```

```{r, echo=T}
train_TS
head(hw_NT_NS$fitted)
```

```{r, echo=T}
predhw_NT_NS = predict(hw_NT_NS, 4, prediction.interval = TRUE)
predhw_NT_NS

forecast(hw_NT_NS, h = 4)

```

##### Evaluate Model

```{r, echo=T}

regr.eval(train$MeanPrice[2:length(train$MeanPrice)], hw_NT_NS$fitted[,1])

regr.eval(test$MeanPrice, predhw_NT_NS[,1])

plot(hw_NT_NS, predhw_NT_NS)

```

##### HoltWinters Model with Trend and no Seasonality  

```{r, echo=T}

hw_T_NS = HoltWinters(train_TS, beta=T, gamma=F)
hw_T_NS

train_TS
head(hw_T_NS$fitted)

predhw_T_NS = predict(hw_T_NS, 4, prediction.interval = TRUE)
head(predhw_T_NS)

```

##### Evaluate Model

```{r, echo=T}

regr.eval(train$MeanPrice[3:length(train$MeanPrice)], hw_T_NS$fitted[,1])

regr.eval(test$MeanPrice, predhw_T_NS[,1])

plot(hw_T_NS, predhw_T_NS)
```

##### HoltWinters model with trend and Seasonality

```{r, echo=T}

hw_T_S = HoltWinters(train_TS, beta=T, gamma=T)
hw_T_S = HoltWinters(train_TS)
hw_T_S

head(train_TS)
head(hw_T_S$fitted)

predhw_T_S = predict(hw_T_S, 4, prediction.interval = TRUE)
head(predhw_T_S)

```

##### Evaluate Model

```{r, echo=T}

regr.eval(train$MeanPrice[13:length(train$MeanPrice)], hw_T_S$fitted[,1])

regr.eval(test$MeanPrice, predhw_T_S[,1])

plot(hw_T_NS, predhw_T_NS)
```


#### Auto-Regressive Models

##### ACF / PACF Plots 

* ACF: n th lag of ACF is the correlation between a day and n days before that.

* PACF: The same as ACF with all intermediate correlations removed.

    * Ideal Trend      : Decreasing ACF and 1 or 2 lags of PACF
    * Ideal Seasonality: Cyclicality in ACF and a few lags of PACF with some positive and some negative
    * Ideal Random     : A spike may or may not be present; even if present, magnitude will be small

```{r, echo=T}

par(mfrow=c(1, 2))

Acf(train_TS, lag = 44)
Pacf(train_TS, lag = 44)

```

##### ACF and PACF for one Differencing

```{r, echo=T}

par(mfrow=c(1, 2))

Acf(diff(train_TS,lag=1), lag=43) 
Pacf(diff(train_TS, lag=1),lag=43)

```

##### Differencing Terms

`ndiffs` and `nsdiffs` functions of forecast package can be used to findout the number of differences and seasonal differences, required to stationarize the data.

```{r, echo=T}

ndiffs(train_TS)
nsdiffs(train_TS)
ndiffs(nsdiffs(train_TS))

```

***
